When easy is not preferred: A discounting paradigm to assess load-independent task preference

### 2023-03-29 06_Paper/COG-ED/Stage 2/Manuscript.Rmd
some final adjustments

---
title             : "When easy is not preferred: A discounting paradigm to assess load-independent task preference"
shorttitle        : "The CAD paradigm to assess task preference"
author: 
  - name          : "Josephine Zerna"
    orcid         : "0000-0003-2892-884X"
    affiliation   : "1"
    corresponding : yes
    address       : "Zellescher Weg 17, 01069 Dresden, Germany"
    email         : "josephine.zerna@tu-dresden.de"
    equal_contrib : yes
    role:         # Contributorship roles (e.g., CRediT, credit.niso.org)
      - Conceptualization
      - Data curation
      - Methodology
      - Funding acquisition
      - Formal analysis
      - Investigation
      - Project administration
      - Software
      - Visualization
      - Writing - original draft
      - Writing - review & editing
  - name          : "Christoph Scheffel"
    orcid         : "0000-0001-5963-9229"
    affiliation   : "1"
    equal_contrib : yes
    role:
      - Conceptualization
      - Methodology
      - Funding acquisition
      - Investigation
      - Project administration
      - Software
      - Writing - review & editing
  - name          : "Corinna Kührt"
    orcid         : "0000-0002-6418-6479"
    affiliation   : "1"
    role:
      - Formal analysis
      - Writing - review & editing
  - name          : "Alexander Strobel"
    orcid         : "0000-0002-9426-5397"
    affiliation   : "1"
    role: 
      - Conceptualization
      - Resources
      - Supervision
      - Funding acquistion
      - Writing - review & editing
affiliation:
  - id            : "1"
    institution   : "Faculty of Psychology, Technische Universität Dresden, 01062 Dresden, Germany"
authornote        :
abstract          : |
  When individuals set goals, they consider the subjective value (SV) of the anticipated reward and the required effort, a trade-off that is of great interest to psychological research.
  One approach to quantify the SVs of levels of difficulty of a cognitive task is the Cognitive Effort Discounting Paradigm by Westbrook and colleagues (2013).
  However, it fails to acknowledge the highly subjective nature of effort, as it assumes a unidirectional, inverse relationship between task load and SVs.
  Therefore, it cannot map differences in effort perception that arise from traits like Need for Cognition, since individuals who enjoy effortful cognitive activities likely do not prefer the easiest level.
  We replicated the analysis of Westbrook and colleagues with an adapted version, the Cognitive and Affective Discounting (CAD) Paradigm.
  It quantifies SVs without assuming that the easiest level is preferred, thereby enabling the quantification of SVs for tasks without objective order of task load.
  Results show that many participants preferred a more or the most difficult level.
  Variance in SVs was best explained by a declining logistic contrast of the $n$-back levels and by the accuracy of responses, while reaction time as a predictor was highly volatile depending on the preprocessing pipeline.
  Participants with higher Need for Cognition scores perceived higher $n$-back levels as less effortful and found them less aversive.
  Effects of Need for Cognition on SVs in lower levels did not reach significance, as group differences only emerged in higher levels.
  The CAD Paradigm appears to be well suited for assessing and analysing task preferences independent of the supposed objective task difficulty.

@@

```{r figure1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="An example for subjective values for an $n$-back task with three levels, returned by different modifications of the COG-ED paradigm for a hypothetical participant with the true preference 2-back > 3-back > 1-back. The grey boxes are the choice options shown to the participant. The participant's final reward value of the flexible level is displayed after the first arrow. The resulting subjective value of each level is displayed after the second arrow, in the notation \"SV 3-back(1-back)\" for the subjective value of 3-back when 1-back is the other choice. The Solution and Additional Benefit panel follow the same logic, but are preceded by a choice between equal rewards, and the participant's first choice indicated by an exclamation mark. Figure available at https://osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

For each rmANOVA we report the generalized eta squared $\\hat{\\eta}^2_G$, which estimates the effect size in analyses that contain both manipulated and non-manipulated terms.
We used rmANOVAs to investigate whether objective performance measures and subjective task load measures changed across $n$-back levels.
The performance measure $d'$ did not change across $n$-back levels (`r apa_print(hypothesis1a_rmanova)$full_result$level`, `r apa_print(hypothesis1a_BF)$statistic`), but the median RT did (`r apa_print(hypothesis1b_rmanova)$full_result$level`, `r apa_print(hypothesis1b_BF)$statistic`), evidence was not in favour of H1a but in favour of H1b.
Specifically, the median RT was higher for the more difficult level in every contrast, with two exceptions:
It did not differ between 2- and 4-back, and it was higher for 3- than for 4-back (Table\ \@ref(tab:H1b-contrasts)).

@@

Consequently, we had to adapt the MLM to incorporate this non-linear trend.
To apply the contrast to the $n$-back levels, we had to turn the variables into a factor, with two consequences:
Centered variables cannot be turned into factors, so we entered the variable level in its raw form, and factors cannot be used as random slopes, so the model is now defined as:
$$
SV \sim level\ + d' + median RT + (1|subject)
$$
This means that the intercept still varied between subjects, but there were no random slopes anymore.
To provide more than one observation per factor level, we used the two rounds per $n$-back level per subject, rather than $n$-back levels per subject.
The ICC of the null model indicated that there was a correlation of $r=$ `r sub("^(-?)0.", "\\1.", sprintf("%.2f", round(model0_h2b_ICC, digits = 3)))` between the SVs of a subject, i.e. that `r paste0(formatC(model0_h2b_ICC * 100, format = "f", digits = 2), "%")` of variance in SVs could be explained by differences between participants.
We did not use an optimization algorithm to improve the fit of the random intercept model.
A total of `r nrow(excl_cases_h2b)` data points from `r length(unique(excl_cases_h2b$subject))` participants were excluded, because the residuals exceeded 3 SD above the mean.
The results of the final model are displayed in Table\ \@ref(tab:H2b-results).

@@

```{r sca-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Results of the multi level model for each of the 63 preprocessing pipelines. Drawing a vertical through both panels indicates the type of preprocessing (panel b) of the pipeline and the resulting beta estimates of the three predictors in the model (panel a). The colourbar in panel b indicates the BF10 of each multi level model compared to a model in which the n-back level has no effect. The pipelines in both panels are sorted left to right in descending order of the magnitude of the beta estimate of the predictor n-back level. Figure available at https://osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

```{r nfcgroups-sv, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Subjective values per n-back level for participants with Need for Cognition (NFC) scores above and below the median. $N=116$. The scatter has a horizontal jitter of 0.2. Smoothing of conditional means with Loess method. Figure available at https://osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

```{r h3ac-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Difference scores for subjective values (a) and aversiveness ratings (b) when subtracting 2- from 1-back and 3- from 2-back. Horizontal lines of the boxplots represent the median per group, whiskers represent 1.5 interquartile ranges. NFC = Need for Cognition score. $N=116$. Figure available at https://osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

The rmANOVA on the association between NFC scores and NASA-TLX scores revealed a main effect of $n$-back level (`r apa_print(hypothesis3b_model)$full_result$level`, `r apa_print(hypothesis3b_BF_levels)$statistic$level`) and an interaction between $n$-back level and NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian_level`), but no main effect of NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian`, `r apa_print(hypothesis3b_BF_levels)$statistic$nfcmedian`).
Post-hoc tests showed that the participants with NFC scores below the median had higher NASA-TLX scores for 3-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X3_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 3], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 3], progress = FALSE, paired = FALSE))$statistic`) and for 4-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X4_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 4], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 4], progress = FALSE, paired = FALSE))$statistic`) than those with NFC scores above the median, so evidence was in favour of H3b.
Regardless of NFC scores, NASA-TLX scores were higher for the more difficult level in each pair of $n$-back levels (Figure\ \@ref(fig:h3b-plot)).

```{r h3b-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="NASA-TLX sum scores for each $n$-back level. Colours indicate Need for Cognition (NFC) score above or below the median. $N=116$. Figure available at https://osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

In the analysis of SV difference scores, the NFC group did not reach significance as a predictor.
This was likely due to the bandwidth of SVs of participants with NFC scores around the median, and due to the fact that the difference appeared most pronounced for 4-back, and we only analyzed the difference scores between 1- and 2-back and 2- and 3-back.
As the exploratory analysis showed, only 4-back yielded a significant group difference, and SVs of participants with NFC scores above the median were higher for 2- to 4-back and lower for 1-back.
The analysis of NASA-TLX scores showed that the sum score increased with every $n$-back level, and that participants with NFC scores below the median had higher NASA-TLX scores for 3- and 4-back than those below the median.
This demonstrates that higher $n$-back levels have a higher discriminatory power regarding inter-individual differences in subjective effort perception.
This was also supported by the fact that higher $n$-back levels were perceived as more aversive, and participants with NFC scores below the median reported higher aversion than those with NFC scores above the median.
Our data supports the notion of a nonlinear interaction between person and situation that has also been described by Schmidt et al. (2013)[-@Schmidt2013] and Blum et al. (2018)[-@Blum2018] in the same-named NIPS model.
The NIPS model describes behaviour as a function of situational affordance which is mediated by personality traits.
The behavioural variability follows an s-shaped curve, such that "strong" situations with low or high situational affordance elicit the least behavioural variability, while "weak" situations with moderate affordance maximize individual differences.
These differences are caused by a person's expression of a certain trait, which shifts the curve along the y-axis.
In our study, the situational affordance is the $n$-back level and the behaviour is the SV, following a declining logistic curve, i.e. a mirrored s-shape.
Hence, the variability in SVs increased from 1- to 4-back, and participants with higher NFC showed a more shallow decline in SVs as the situational affordance approached moderate values.
According to the NIPS model, we can expect the SVs of participants with higher and lower NFC to converge again in levels of $n>4$, since behavioural variability decreases when situational affordance is high.
An investigation of this relationship using the COG-ED paradigm[-@Westbrook2013] had been encouraged by Strobel et al.[@Strobel2020] based on their findings on demand avoidance and cognitive effort investment.
With the CAD paradigm, the declining logistic contrast of SVs across levels resembles the ascending logistic curve of the NIPS model[@Schmidt2013;@Blum2018] and should be tested further in a setting with $n$-back levels exceeding $n=4$.

>>>>

---
title             : "When easy is not preferred: A discounting paradigm to assess load-independent task preference"
shorttitle        : "The CAD paradigm to assess task preference"
author: 
  - name          : "Josephine Zerna"
    orcid         : "0000-0003-2892-884X"
    affiliation   : "1"
    corresponding : yes
    address       : "Zellescher Weg 17, 01069 Dresden, Germany"
    email         : "josephine.zerna@tu-dresden.de"
    equal_contrib : yes
    role:         # Contributorship roles (e.g., CRediT, credit.niso.org)
      - Conceptualization
      - Data curation
      - Methodology
      - Funding acquisition
      - Formal analysis
      - Investigation
      - Project administration
      - Software
      - Visualization
      - Writing - original draft
      - Writing - review & editing
  - name          : "Christoph Scheffel"
    orcid         : "0000-0001-5963-9229"
    affiliation   : "1"
    equal_contrib : yes
    role:
      - Conceptualization
      - Methodology
      - Funding acquisition
      - Investigation
      - Project administration
      - Software
      - Writing - review & editing
  - name          : "Corinna Kührt"
    orcid         : "0000-0002-6418-6479"
    affiliation   : "1"
    role:
      - Formal analysis
      - Writing - review & editing
  - name          : "Alexander Strobel"
    orcid         : "0000-0002-9426-5397"
    affiliation   : "1"
    role: 
      - Conceptualization
      - Resources
      - Supervision
      - Funding acquistion
      - Writing - review & editing
affiliation:
  - id            : "1"
    institution   : "Faculty of Psychology, Technische Universität Dresden, 01062 Dresden, Germany"
authornote        :
abstract          : |
  When individuals set goals, they consider the subjective value (SV) of the anticipated reward and the required effort, a trade-off that is of great interest to psychological research.
  One approach to quantify the SVs of levels of difficulty of a cognitive task is the Cognitive Effort Discounting Paradigm by Westbrook and colleagues (2013).
  However, it fails to acknowledge the highly individual nature of effort, as it assumes a unidirectional, inverse relationship between task load and SVs.
  Therefore, it cannot map differences in effort perception that arise from traits like Need for Cognition, since individuals who enjoy effortful cognitive activities likely do not prefer the easiest level.
  We replicated the analysis of Westbrook and colleagues with an adapted version, the Cognitive and Affective Discounting (CAD) Paradigm.
  It quantifies SVs without assuming that the easiest level is preferred, thereby enabling the assessment of SVs for tasks without objective order of task load.
  Results show that many participants preferred a more or the most difficult level.
  Variance in SVs was best explained by a declining logistic contrast of the $n$-back levels and by the accuracy of responses, while reaction time as a predictor was highly volatile depending on the preprocessing pipeline.
  Participants with higher Need for Cognition scores perceived higher $n$-back levels as less effortful and found them less aversive.
  Effects of Need for Cognition on SVs in lower levels did not reach significance, as group differences only emerged in higher levels.
  The CAD Paradigm appears to be well suited for assessing and analysing task preferences independent of the supposed objective task difficulty.

@@

```{r figure1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="An example for subjective values for an $n$-back task with three levels, returned by different modifications of the COG-ED paradigm for a hypothetical participant with the true preference 2-back > 3-back > 1-back. The grey boxes are the choice options shown to the participant. The participant's final reward value of the flexible level is displayed after the first arrow. The resulting subjective value of each level is displayed after the second arrow, in the notation \"SV 3-back(1-back)\" for the subjective value of 3-back when 1-back is the other choice. The Solution and Additional Benefit panel follow the same logic, but are preceded by a choice between equal rewards, and the participant's first choice indicated by an exclamation mark. Figure available at osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

We used rmANOVAs to investigate whether objective performance measures and subjective task load measures changed across $n$-back levels.
For each rmANOVA we report the generalized eta squared $\hat{\eta}^2_G$, which estimates the effect size in analyses that contain both manipulated and non-manipulated terms.
The performance measure $d'$ did not change across $n$-back levels (`r apa_print(hypothesis1a_rmanova)$full_result$level`, `r apa_print(hypothesis1a_BF)$statistic`), but the median RT did (`r apa_print(hypothesis1b_rmanova)$full_result$level`, `r apa_print(hypothesis1b_BF)$statistic`), evidence was not in favour of H1a but in favour of H1b.
Specifically, the median RT was higher for the more difficult level in every contrast, with two exceptions:
It did not differ between 2- and 4-back, and it was higher for 3- than for 4-back (Table\ \@ref(tab:H1b-contrasts)).

@@

Consequently, we had to adapt the MLM to incorporate this non-linear trend.
To apply the contrast to the $n$-back levels, we had to turn the variables into a factor, with two consequences:
Centered variables cannot be turned into factors, so we entered the variable level in its raw form, and factors cannot be used as random slopes, so the model is now defined as:
$$
SV \sim level\ + d' + median RT + (1|subject)
$$
This means that the intercept still varied between subjects, but there were no random slopes anymore.
To provide more than one observation per factor level, we used the two rounds per $n$-back level per subject, rather than $n$-back levels per subject.
The ICC of the null model indicated that there was a correlation of $r=$ `r sub("^(-?)0.", "\\1.", sprintf("%.3f", round(model0_h2b_ICC, digits = 3)))` between the SVs of a subject, i.e. that `r paste0(formatC(model0_h2b_ICC * 100, format = "f", digits = 2), "%")` of variance in SVs could be explained by differences between participants.
We did not use an optimization algorithm to improve the fit of the random intercept model.
A total of `r nrow(excl_cases_h2b)` data points from `r length(unique(excl_cases_h2b$subject))` participants were excluded, because the residuals exceeded 3 SD above the mean.
The results of the final model are displayed in Table\ \@ref(tab:H2b-results).

@@

```{r sca-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Results of the multi level model for each of the 63 preprocessing pipelines. Drawing a vertical through both panels indicates the type of preprocessing (panel b) of the pipeline and the resulting beta estimates of the three predictors in the model (panel a). The colourbar in panel b indicates the BF10 of each multi level model compared to a model in which the n-back level has no effect. The pipelines in both panels are sorted left to right in descending order of the magnitude of the beta estimate of the predictor n-back level. Figure available at osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

```{r nfcgroups-sv, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Subjective values per n-back level for participants with Need for Cognition (NFC) scores above and below the median. $N=116$. The scatter has a horizontal jitter of 0.2. Smoothing of conditional means with Loess method. Figure available at osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

```{r h3ac-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Difference scores for subjective values (a) and aversiveness ratings (b) when subtracting 2- from 1-back and 3- from 2-back. Horizontal lines of the boxplots represent the median per group, whiskers represent 1.5 interquartile ranges. NFC = Need for Cognition score. $N=116$. Figure available at osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

The rmANOVA on the association between NFC scores and NASA-TLX scores revealed a main effect of $n$-back level (`r apa_print(hypothesis3b_model)$full_result$level`, `r paste(unlist(strsplit(apa_print(hypothesis3b_BF_levels)$statistic$level, split='} \\pm', fixed=TRUE))[1],"}$")`) and an interaction between $n$-back level and NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian_level`), but no main effect of NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian`, `r paste(unlist(strsplit(apa_print(hypothesis3b_BF_levels)$statistic$nfcmedian, split='} \\pm', fixed=TRUE))[1],"}$")`).
Post-hoc tests showed that the participants with NFC scores below the median had higher NASA-TLX scores for 3-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X3_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 3], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 3], progress = FALSE, paired = FALSE))$statistic`) and for 4-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X4_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 4], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 4], progress = FALSE, paired = FALSE))$statistic`) than those with NFC scores above the median, so evidence was in favour of H3b.
Regardless of NFC scores, NASA-TLX scores were higher for the more difficult level in each pair of $n$-back levels (Figure\ \@ref(fig:h3b-plot)).

```{r h3b-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="NASA-TLX sum scores for each $n$-back level. Colours indicate Need for Cognition (NFC) score above or below the median. $N=116$. Figure available at osf.io/vnj8x/, under a CC-BY-4.0 license.", fig.pos="H"}

@@

In the analysis of SV difference scores, the NFC group did not reach significance as a predictor.
This was likely due to the bandwidth of SVs of participants with NFC scores around the median, and due to the fact that the difference appeared most pronounced for 4-back, and we only analyzed the difference scores between 1- and 2-back and 2- and 3-back.
As the exploratory analysis showed, only 4-back yielded a significant group difference, and SVs of participants with NFC scores above the median were higher for 2- to 4-back and lower for 1-back.
The analysis of NASA-TLX scores showed that the sum score increased with every $n$-back level, and that participants with NFC scores below the median had higher NASA-TLX scores for 3- and 4-back than those below the median.
This demonstrates that higher $n$-back levels have a higher discriminatory power regarding inter-individual differences in subjective effort perception.
This was also supported by the fact that higher $n$-back levels were perceived as more aversive, and participants with NFC scores below the median reported higher aversion than those with NFC scores above the median.
Our data supports the notion of a nonlinear interaction between person and situation that has also been described by Schmitt et al. (2013)[-@Schmitt2013] and Blum et al. (2018)[-@Blum2018] in the same-named NIPS model.
The NIPS model describes behaviour as a function of situational affordance which is mediated by personality traits.
The behavioural variability follows an s-shaped curve, such that "strong" situations with low or high situational affordance elicit the least behavioural variability, while "weak" situations with moderate affordance maximize individual differences.
These differences are caused by a person's expression of a certain trait, which shifts the curve along the y-axis.
In our study, the situational affordance is the $n$-back level and the behaviour is the SV, following a declining logistic curve, i.e. a mirrored s-shape.
Hence, the variability in SVs increased from 1- to 4-back, and participants with higher NFC showed a more shallow decline in SVs as the situational affordance approached moderate values.
According to the NIPS model, we can expect the SVs of participants with higher and lower NFC to converge again in levels of $n>4$, since behavioural variability decreases when situational affordance is high.
An investigation of this relationship using the COG-ED paradigm[-@Westbrook2013] had been encouraged by Strobel et al.[@Strobel2020] based on their findings on demand avoidance and cognitive effort investment.
With the CAD paradigm, the declining logistic contrast of SVs across levels resembles the ascending logistic curve of the NIPS model[@Schmitt2013;@Blum2018] and should be tested further in a setting with $n$-back levels exceeding $n=4$.
